---
title: "Reproducible Research - Self Monitor Activity Dataset Analysis"
author: "My D. Coyne"
date: "March 9, 2015"
output:
  html_document:
    keep_md: true
---

    
---

### Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. In this assignment one will analyze a self monintor dataset;  the data is collected by an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  **The objective of the assignment is to show how tool such as knitr and R-mardow would help analysts in creating reproducible data analysis research project.**

### About the Dataset

The variables included in this dataset are:

    * **steps:**  Number of steps taking in a 5-minute interval (missing values are coded as NA)

    * **date:** The date on which the measurement was taken in YYYY-MM-DD format

    * **interval:**  Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

The activity data set is read in memory using read.table(), with comma is a separator and convert to a data frame.  Some prelimiary observations on the dataset:

1.  There are total of 17,568 observation; after removing the obseravations that contain NAs, there are 15,264 observations remained.

2.  The observations spanned through 61 days; of which only 53 days that contains non-NAs observations.  


```{r Load_initEx}

dat <-  data.frame(read.table
                   ("~/Documents/Coursera/ReproR/GitTest/RepData_PeerAssessment1/activity.csv"
                    , header=TRUE, sep=","))

## Removing NA's
df <- na.omit(dat);
nrow(df)

## Number of days spanned with NA's
tDays <- unique(dat$date)
length(tDays)

## Number of days spanned without NA's
t <- unique(df$date)
length(t)
```


## What is mean total number of steps taken per day?

In this section, the observations with missing values are ignored, or not included in the analysis.  

1.  Calculate the total number of steps taken per day.  The rows in the dataset can be grouped into each day, then one can sum up the steps of the observed intervals of the day.  To do this, one can use aggregate() function -- aggreage data base on date, and apply summaation on the steps, as in the below line of code: 

```{r}
totalStepsPerDay <-aggregate(x=df$steps,list(date=df$date), FUN=sum, na.rm=TRUE);
totalStepsPerDay
```

2.  Histogram to show totals steps taken each day

```{r hist_total_steps}
library (ggplot2)

h <-ggplot(data=totalStepsPerDay, aes(x=x))+
          ggtitle("Histogram of total steps per day (binwidth=500)") + 
          xlab("Total Steps")
h + geom_histogram(binwidth=500, aes(fill = ..count..)) +
    scale_fill_gradient("Count", low="#fccde5", high="#ae017e")

```



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

